Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-31 main.Rmd
data structure image update: include all cyclic granularities

Effective exploration or good visualization require good data structures. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of linear and cyclic granularities. A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in (Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities. Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\} and $v$ denotes the measurement variable. Data sets of the form <$C_1$, $C_2$, $v$> can then form the basis for exploration and analysis of the measured variable across bivariate cyclic granularities.

@@

```{r data-structure, fig.cap = "The data structure for exploring periodicities in data by including two cyclic granularities in the tsibble structure with index, key and measured variables.", fig.pos="htb"}
include_graphics("Figs/data-structure-v1.png")
```

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). While considering data sets of the form $<C_1, C_2, v>$, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets for each combination of levels ($A_i$, $B_j$) like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$  which will lead to $nm$ data subsets. 
Now, some situations can lead to few or many of these sets being empty. We will discuss few cases, where one or more of these $nm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->

>>>>

Effective exploration or good visualization require good data structures. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of linear and cyclic granularities. A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities $C_i$ such that $C_i$ maps index set to $n_i$ levels or categories and $v$ denotes the measurement variable. The total number of cyclic granularities would be based on the number of linear granularities considered in the hierarchy table and presence of any aperiodic cyclic granularities. Suppose we have $n$ linear granularities in the hierarchy table, $n(n-1)/2$ circular or quasi-circular cyclic granularities each relating one finer linear granularity with a coarser one could be constructed. Any attempt to encode all or many cyclic granularities at the same time to develop insights on periodicity might fail or become clumsy. Instead, this big problem is broken down by focusing on two cyclic granularities at a time. Hence data sets of the form <$C_i$, $C_j$, $v$> forms the basis for exploration and analysis of the measured variable.

<!-- CONTENT SHORT -->
<!-- across bivariate cyclic granularities.  -->
<!-- Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\} and $v$ denotes the measurement variable. Data sets of the form <$C_1$, $C_2$, $v$> can then form the basis for exploration and analysis of the measured variable across bivariate cyclic granularities.  -->

@@

```{r data-structure, fig.cap = "The data structure for exploring periodicities in data by including cyclic granularities in the tsibble structure with index, key and measured variables.", fig.pos="htb"}
include_graphics("Figs/data-structure-v2.png")
```

@@

<!-- Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\} and $v$ denotes the measurement variable. Data sets of the form <$C_1$, $C_2$, $v$> can then form the basis for exploration and analysis of the measured variable across bivariate cyclic granularities. -->

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_l$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\}, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets for each combination of levels ($A_i$, $B_j$) like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, l}$ and $j \in {1, 2, \dots, m}$  which will lead to $lm$ data subsets. 
Now, some situations can lead to few or many of these sets being empty. We will discuss few cases, where one or more of these $lm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->
